Graded Call-By-Push-Value

### 2020-07-07 popl21/popl21.tex
[ popl21 ] denotational coeffect grading

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}
\newcommand{\Den}[2]{\den{#1}_{#2}}

\newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\Eff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\bu}{\ensuremath{\bullet}}
\newcommand{\Ge}{\ensuremath{\varepsilon}}
\newcommand{\Ga}{\ensuremath{\alpha}}
\newcommand{\Gd}{\ensuremath{\delta}}
\newcommand{\Gl}{\ensuremath{\lambda}}
\newcommand{\Gr}{\ensuremath{\rho}}
\newcommand{\GG}{\ensuremath{\Gamma}}
\newcommand{\GS}{\ensuremath{\Sigma}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\I}{\mathsf{I}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\wleq}{\sqsubseteq}
\newcommand{\winf}{\sqcap}
\newcommand{\wsup}{\sqcup}
%\newcommand{\winf}{\bigsqcap}
\newcommand{\emp}{0_{\W}}
% \newcommand{\emp}{\varepsilon}
% \newcommand{\emp}{\emptyset}

\newcommand{\CC}{\ensuremath{\mathcal{C}}}
\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
\newcommand{\SET}{\mathsf{SET}}
\newcommand{\CPO}{\mathsf{CPO}}
\newcommand{\T}[1][]{\mathsf{T}_{#1}}
\newcommand{\D}[1][]{\mathsf{D}_{#1}}
\newcommand{\trun}{\mathsf{run}}
\newcommand{\run}[1][]{\mathsf{run}_{#1}}
\newcommand{\texpose}{\mathsf{expose}}
\newcommand{\expose}[1][]{\texpose_{#1}}
\newcommand{\tA}{\mathsf{A}}
\newcommand{\tB}{\mathsf{B}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\tExc}{\mathsf{Exc}}
\newcommand{\Exc}{\mathsf{Exc}}
\newcommand{\String}{\mathsf{String}}
\newcommand{\StringLe}[1]{\mathsf{String}{\leq}\,#1}
\newcommand{\tfmap}{\mathsf{fmap}}
\newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
\newcommand{\treturn}{\mathsf{return}}
\newcommand{\tjoin}{\mathsf{join}}
\newcommand{\join}{\mathsf{join}}
\newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
\newcommand{\tcast}{\mathsf{cast}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\tid}{\mathsf{id}}
\newcommand{\teval}{\mathsf{eval}}
\newcommand{\tstrength}{\mathsf{strength}}
\newcommand{\strengthl}{\tstrength^{\tl}}
\newcommand{\strengthr}{\tstrength^{\tr}}
\newcommand{\tl}{\mathsf{l}}
\newcommand{\tr}{\mathsf{r}}
\newcommand{\tcurry}{\mathsf{curry}}
\newcommand{\comp}{\circ}
\newcommand{\tweaken}{\mathsf{weaken}}
\newcommand{\tdrop}[1][]{\mathsf{drop}_{#1}}
%\newcommand{\drop}[1][]{\tdrop_{#1}} % Name clash
\newcommand{\tcontract}{\mathsf{contract}}
\newcommand{\contract}[1][]{\tcontract_{#1}}

@@

\newcommand{\sumt}   [2]{\ensuremath{\Sigma_{#1} #2}}
\newcommand{\tupt}   [2]{\ensuremath{\otimes_{#1} #2}}
\newcommand{\rect}   [2]{\ensuremath{\Pi_{#1} #2}}
\newcommand{\sumty}  [3]{\ensuremath{\Sigma_{#1:#2} #3_{#1}}}
\newcommand{\tupty}  [3]{\ensuremath{\otimes_{#1:#2} #3_{#1}}}
\newcommand{\recty}  [3]{\ensuremath{\Pi_{#1:#2} #3_{#1}}}
\newcommand{\thunkty}[2]{\ensuremath{[#1]#2}}
\newcommand{\compty} [2]{\ensuremath{\langle#1\rangle#2}}

@@

\subsection{Effect-graded CBPV: syntax and typing}

@@

\citet{mcBride:wadler60} observed that resource contexts
$\gamma : \R^I$ form a left $\R$-semimodule under pointwise addition
$(\gamma + \delta)(i) = \gamma(i) + \delta(i)$ and scaling
$(r \cdot \gamma)(i) = r \cdot \gamma(i)$.  Contexts $\Gamma$ can thus
be interpreted as $R^{\dom \Gamma}$-graded $\D$-coalgebras $C$ with
the following operations:
\[
\begin{array}{lcl}
  \tdrop[C]     & : & \Hom {C_0} {\I} \\
  \duplicate[C] & : & \Hom {C_{\gamma+\delta}} {C_\gamma \otimes C_\delta} \\
  \duplicate[C] & : & \Hom {C_{\sumt I \gamma}} {\tupt {i:I} C_{\gamma_i}} \\
  \expose[C]    & : & \Hom {C_{r\gamma}} {\D[r]\,C_\gamma} \\
\end{array}
\]
The second, generalized form of $\tduplicate$ will be used to split
resources accumulated from $I$ parties.

We will also need to interpret context extension.  To this end, we
shall employ a structured product
$(C \boxtimes D)_{\gamma.\delta} = C_\gamma \otimes D_\delta$ where
$\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitely use the
isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
and $\R$-graded $\D$-coalgebras $A : \R \to \CC$ and define
\[
\den{\ext \Gamma x P}
  = \den \Gamma \boxtimes \denp P
\]
sweeping name issues under the carpet (to be handled by de Bruijn indices).

@@

  & ::= & \diamond P % \graybox{\compty r P}

@@

         \ret v        \mid \qbind {\graybox r} x u t

@@

     {\qCompTy \gamma \Gamma u N \qquad
      \CompTy {\qext \delta \Gamma r y N} t {N'}
    }{\gpCompTy {r\gamma + \delta} \Gamma {\qlet r y u t} {N'}}

@@

     {\gqValTy \gamma \Gamma v P \qquad
      \CompTy {\gqext \delta \Gamma r x P} t N
    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqlet r x v t} N}

@@

     {\gqValTy \gamma \Gamma v {\Box N} \qquad
      \CompTy {\gqext \delta \Gamma r y N} t {N'}
    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqforced r v y t} {N'}}

@@

     {\gqValTy \gamma \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\gqext \delta \Gamma r {x_i} {P_i}} {t_i} N
    }{\gpCompTy {r\gamma + \delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}

@@

     {\gqValTy \gamma \Gamma v {\tupt I P} \qquad

@@

    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqsplits r v {\bar x} t} N}

@@

    }{\qCompTy \gamma \Gamma {\ret v} {\diamond P}}

@@

     {\gqCompTy \gamma \Gamma u {\diamond P} \qquad
      \CompTy {\gqext \delta \Gamma r x P} t  N
    }{\gpCompTy {r\gamma + \delta} \Gamma {\gqbind r x u t} N}

@@

\subsection{Coeffect-graded CBPV: typing and semantics}

We now have the mathematical structures in place to define a
coeffect-graded variant of CBPV (see Fig.~\ref{coeff-cbpv}).
The difference to pure CBPV is laid out in \emph{gray boxes}.

The fundamental novelty is that the typing judgements
$\qValTy \gamma \Gamma v P$ and $\qCompTy \gamma \Gamma t N$ are
equipped with resource contexts $\gamma$ matching the typing contexts
$\Gamma$.  A common pattern in the rules is that resource requirements
of the subterms are added when both subterms are or maybe evaluated at
runtime (rules \rintro\otimes, $\rlet^{+/-}$,
\relim{\Box/\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
the branches $t_i$ of the case statement (\relim\GS) share a resource
context $\delta$, since only one of the branches is executed at
runtime.  The different resource requirements of the branches can be
subsumed under their maximum using rule \rweak.  Similarly, the
components $t_i$ of a record (\rintro\Pi) share a resource context,
since projection retrieves a single component (\relim\Pi).  This is
the opposite of eager tuples (\rintro\otimes) where their eliminations
makes all components available at the same time (\relim\otimes), thus,
all have to be evaluated at runtime.

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  In quantitative type with $\R = \Pot \bN$, qualifier
$r$ gives the possible usage quantities of the function argument,
e.g. $\{1\}$ for exactly one use (linear), $\{0,1\}$ for at most one
use (affine), $\{0\}$ for no use (constant), and $\bN$ for arbitrary
use (unrestricted).  A resource qualifier could also be a security
level (public or private), and many other instances of $\R$ exist.  If
a lambda abstraction $\lam x t$ is typed with $\qfun r P N$, qualifier
$r$ is attached to variable $x$ in the resource context (rule
\rintro\lolli).  If a function $t : \qfun r P N$ is applied (rule
\relim\lolli), we need an $r$-qualified argument $v : rP$.  We could
have given a qualified value typing judgement
$\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allow to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling into
the elimination rules instead.  Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (\relim\Box, \relim\GS, \relim\otimes) and into the
\rlet-rules.

@@

@@

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

>>>>

\newcommand{\denp}{\den[+]}
\newcommand{\denn}{\den[-]}
\newcommand{\Denp}[2]{\denp{#1}_{#2}}
\newcommand{\Denn}[2]{\denn{#1}_{#2}}
\newcommand{\Den}[2]{\den{#1}_{#2}}
\newcommand{\denv}[2]{\dent{#1}_{#2}}

\newcommand{\tEff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\Eff}{\ensuremath{\mathsf{Eff}}}
\newcommand{\bu}{\ensuremath{\bullet}}
\newcommand{\Ge}{\ensuremath{\varepsilon}}
\newcommand{\Ga}{\ensuremath{\alpha}}
\newcommand{\Gd}{\ensuremath{\delta}}
\newcommand{\Gg}{\ensuremath{\gamma}}
\newcommand{\Gl}{\ensuremath{\lambda}}
\newcommand{\Gr}{\ensuremath{\rho}}
\newcommand{\GG}{\ensuremath{\Gamma}}
\newcommand{\GS}{\ensuremath{\Sigma}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\I}{\mathsf{I}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\wleq}{\sqsubseteq}
\newcommand{\winf}{\sqcap}
\newcommand{\wsup}{\sqcup}
%\newcommand{\winf}{\bigsqcap}
\newcommand{\emp}{0_{\W}}
% \newcommand{\emp}{\varepsilon}
% \newcommand{\emp}{\emptyset}

\newcommand{\CC}{\ensuremath{\mathcal{C}}}
\newcommand{\Hom}[2]{\CC\,(#1,\,#2)}
\newcommand{\SET}{\mathsf{SET}}
\newcommand{\CPO}{\mathsf{CPO}}
\newcommand{\T}[1][]{\mathsf{T}_{#1}}
\newcommand{\D}[1][]{\mathsf{D}_{#1}}
\newcommand{\trun}{\mathsf{run}}
\newcommand{\run}[1][]{\mathsf{run}_{#1}}
\newcommand{\texpose}{\mathsf{expose}}
\newcommand{\expose}[1][]{\texpose_{#1}}
\newcommand{\tA}{\mathsf{A}}
\newcommand{\tB}{\mathsf{B}}
\newcommand{\A}{\mathsf{A}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\tExc}{\mathsf{Exc}}
\newcommand{\Exc}{\mathsf{Exc}}
\newcommand{\String}{\mathsf{String}}
\newcommand{\StringLe}[1]{\mathsf{String}{\leq}\,#1}
\newcommand{\tfmap}{\mathsf{fmap}}
\newcommand{\fmap}[2][]{\tfmap_{#1}\,#2}
\newcommand{\fmapt}[2][]{\tfmap_{#1}#2}
\newcommand{\treturn}{\mathsf{return}}
\newcommand{\tjoin}{\mathsf{join}}
\newcommand{\join}{\mathsf{join}}
\newcommand{\append}{\ensuremath{\mathbin{\mathtt{++}}}}
\newcommand{\tcast}{\mathsf{cast}}
\newcommand{\tdist}{\mathsf{dist}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\tid}[1][]{\mathsf{id}_{#1}}
\newcommand{\teval}{\mathsf{eval}}
\newcommand{\tstrength}{\mathsf{strength}}
\newcommand{\strengthl}{\tstrength^{\tl}}
\newcommand{\strengthr}{\tstrength^{\tr}}
\newcommand{\tl}{\mathsf{l}}
\newcommand{\tr}{\mathsf{r}}
\newcommand{\tcurry}{\mathsf{curry}}
\newcommand{\comp}{\circ}
\newcommand{\tweaken}{\mathsf{weaken}}
\newcommand{\tdrop}[1][]{\mathsf{drop}_{#1}}
%\newcommand{\drop}[1][]{\tdrop_{#1}} % Name clash
\newcommand{\tcontract}{\mathsf{contract}}
\newcommand{\contract}[1][]{\tcontract_{#1}}

@@

\newcommand{\sumt}   [2]{\ensuremath{\Sigma_{#1} #2}}
\newcommand{\tupt}   [2]{\ensuremath{\mathop\otimes_{#1} #2}}
\newcommand{\rect}   [2]{\ensuremath{\Pi_{#1} #2}}
\newcommand{\sumty}  [3]{\ensuremath{\Sigma_{#1:#2} #3_{#1}}}
\newcommand{\tupty}  [3]{\ensuremath{\mathop\otimes_{#1:#2} #3_{#1}}}
\newcommand{\recty}  [3]{\ensuremath{\Pi_{#1:#2} #3_{#1}}}
\newcommand{\thunkty}[2]{\ensuremath{[#1]#2}}
\newcommand{\compty} [2]{\ensuremath{\langle#1\rangle#2}}

@@

\subsection{Effect-graded CBPV: syntax and typing}
\label{sec:effect-cbpv}

@@

\citet{mcBride:wadler60} observed that resource contexts
$\gamma : \R^I$ form a left $\R$-semimodule under pointwise addition
$(\gamma + \delta)(i) = \gamma(i) + \delta(i)$ and scaling
$(r \cdot \gamma)(i) = r \cdot \gamma(i)$.  Contexts $\Gamma$ can thus
be interpreted as \emph{$R^{\dom \Gamma}$-graded}
(or \emph{structured}) $\D$-coalgebras $C$ with
the following operations:
\[
\begin{array}{lcl}
  \tdrop[C]     & : & \Hom {C_0} {\I} \\
  \duplicate[C] & : & \Hom {C_{\gamma+\delta}} {C_\gamma \otimes C_\delta} \\
  \duplicate[C] & : & \Hom {C_{\sumt I \gamma}} {\tupt {i:I} C_{\gamma_i}} \\
  \expose[C]    & : & \Hom {C_{r\gamma}} {\D[r]\,C_\gamma} \\
\end{array}
\]
The second, generalized form of $\tduplicate$ will be used to split
resources accumulated from $I$ parties.  Note that
$A_r := C_{r\gamma}$ flattens a structured $\D$-coalgebra $C$ into an
ordinary one ($A$).

We will also need to interpret context extension.  To this end, we
shall employ a structured product
$(C \boxtimes D)_{\gamma.\delta} = C_\gamma \otimes D_\delta$ where
$\gamma : \R^I$ and $\delta : \R^J$ and thus $C \boxtimes D$ is a
$\R^{I+J}$-graded $\D$-coalgebra.  Further, we implicitely use the
isomorphism between $1$-structured $\D$-coalgebras $C : \R^1 \to \CC$
and $\R$-graded $\D$-coalgebras $A : \R \to \CC$ and define
\[
\den{\ext \Gamma x P}
  = \den \Gamma \boxtimes \denp P
  ,
\]
sweeping name issues under the carpet (to be handled by de Bruijn indices).

@@

  & ::= & %\diamond P %
          \graybox{\compty r P}

@@

         \ret v
    \mid \bind x u t
    % \mid \qbind {\graybox r} x u t

@@

     {\qCompTy \delta \Gamma u N \qquad
      \CompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}

@@

     {\gqValTy \delta \Gamma v P \qquad
      \CompTy {\gqext \gamma \Gamma r x P} t N
    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqlet r x v t} N}

@@

     {\gqValTy \delta \Gamma v {\Box N} \qquad
      \CompTy {\gqext \gamma \Gamma r y N} t {N'}
    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}

@@

     {\gqValTy \delta \Gamma v {\sumt I P} \qquad
      \forall i{:}I,\ \CompTy {\gqext \gamma \Gamma r {x_i} {P_i}} {t_i} N
    }{\gpCompTy {\gamma + r\delta} \Gamma {\caser v {\{\graybox{r} x_i.\, t_i\}_{i:I}}} N}

@@

     {\gqValTy \delta \Gamma v {\tupt I P} \qquad

@@

    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqsplits r v {\bar x} t} N}

@@

    }{\qCompTy {\graybox r\gamma} \Gamma {\ret v} {\compty {\graybox r} P}}

@@

     {\gqCompTy \delta \Gamma u {\compty {\graybox r} P} \qquad
      \CompTy {\gqext \gamma \Gamma r x P} t  N
    }{\gpCompTy {\gamma + \delta} \Gamma {\bind x u t} N}
%  \nru{\rintro\diamond}
%      {\qValTy \gamma \Gamma v P
%     }{\qCompTy \gamma \Gamma {\ret v} {\diamond P}}
% \qquad
%  \nru{\relim\diamond}
%      {\gqCompTy \gamma \Gamma u {\diamond P} \qquad
%       \CompTy {\gqext \delta \Gamma r x P} t  N
%     }{\gpCompTy {r\gamma + \delta} \Gamma {\gqbind r x u t} N}

@@

\subsection{Coeffect-graded CBPV: syntax and typing}

We now have the mathematical structures in place to define a
coeffect-graded variant of CBPV (see Fig.~\ref{fig:coeff-cbpv}).
The difference to pure CBPV is laid out in \emph{gray boxes}.

The fundamental novelty is that the typing judgements
$\qValTy \gamma \Gamma v P$ and $\qCompTy \gamma \Gamma t N$ are
equipped with resource contexts $\gamma$ matching the typing contexts
$\Gamma$.  A common pattern in the rules is that resource requirements
of the subterms are added when both subterms are or maybe evaluated at
runtime (rules \rintro\otimes, $\rlet^{+/-}$,
\relim{\Box/\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
the branches $t_i$ of the case statement (\relim\GS) share a resource
context $\gamma$, since only one of the branches is executed at
runtime.  Via rule $\rweak$,
the different resource requirements of the branches can be
subsumed under their maximum wrt.\ $\leq_\R$.  Similarly, the
components $t_i$ of a record (\rintro\Pi) share a resource context
since projection only retrieves one of the components (\relim\Pi).  This is
the opposite of eager tuples (\rintro\otimes) where the elimination
makes all components available at the same time (\relim\otimes), thus,
all components have to be evaluated at runtime.

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  (This type is often written $!_{r}P \lolli N$.)  In
quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
possible usage quantities of the function argument, e.g. $\{1\}$ for
exactly one use (linear), $\{0,1\}$ for at most one use (affine),
$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
(unrestricted).  A resource qualifier could also be a security level
(public or private), or a sensitivity level (a non-negative real)
\citep{reedPeirce:icfp10}.  If a lambda abstraction $\lam x t$ is
typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
the resource context (rule \rintro\lolli).  If a function
$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
$r$-qualified argument $v : rP$.  We could have given a qualified
value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allows to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
into the transition from values to computations.
Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
\rlet-rules, and into \rintro\diamond.

The monadic type $\compty r P$ records the multiplicity $r$ of the
value of type $P$ resulting from the computation.  This construction
is dual to the comonadic type $\thunkty e N$ from effect-graded CBPV
(Section~\ref{sec:effect-cbpv}).  Since negative types are not
$\D$-coalgebras and do not support scaling, the scaling in
$\rintro\diamond$ is the last opportunity for scaling before entering
the monad.  Typing of bind ($\bind x u t$) attaches the resource qualifier $r$ store in $\compty r P$ to the variable $x$ (\relim\diamond).

@@

\subsection{Coeffect-graded CBPV: denotational semantics}

%\subsubsection{Interpretation of types}

Positive types $P$ are interpreted as $\D$-algebras
$\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
$\denn N : \CC$, except for in contexts, where they are interpreted as
$\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
the interpretation of contexts is thus
$\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
\[
\begin{array}{lcl}
  \Denp\_\_ & : & \PTy \to [(\R,\leq) \to \CC] \\
  \Denp{\Box N}    r & = & \D[r]\, \denn N \\
  \Denp{\sumt I P} r & = & \coprod_{i:I} \Denp{P_i} r \\
  \Denp{\tupt I P} r & = & \otimes_{i:I} \Denp{P_i} r \\
\end{array}
\qquad\qquad
\begin{array}{lcl}
  \denn\_\_ & : & \NTy \to \CC \\
  \denn{\compty r P} & = & \T \Denp P r \\
  \denn{\qfun r P N} & = & \Denp P r \lolli \denn N \\
  \denn{\rect I N}   & = & \prod_{i:I} \denn{N_i} \\
\end{array}
\]
The symmetric monoidal category $\CC$ needs to come with cartesian
products $\rect I B$ as well as with distributive coproducts
$\coprod_I A$ with
$\tdist : \Hom {C \otimes \coprod_I A} {\coprod_{i:I} {(C \otimes A_i)}}$.

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose{\den \Gamma}
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
\rvar^-
  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
\\
\rlet^{-}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
\\ &&& {~}
  \comp \duplicate
\\
\rlet^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\expose[\denp P] \comp \denv v r))
  \comp \duplicate
\\
\relim\Box
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

@@

% %% Appendix
% \appendix
% \section{Appendix}

% Text of appendix \ldots
